spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data

### 2020-08-24 paper/spyrison-cook.tex
post proofing changes

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\bibliography{spinifex_paper}

@@

Exploring multivariate spaces is a challenging task, increasingly so as
dimensionality increases. Traditionally, static low-dimensional
projections are used to display multivariate data in two dimensions
including principal component analysis, linear discriminant spaces or
projection pursuit. These are useful for finding relationships between
multiple variables, but they are limited because they show only a
glimpse of the high-dimensional space. An alternative approach is to use
a tour (Asimov \protect\hyperlink{ref-asimov_grand_1985}{1985}) of
dynamic linear projections to look at many different low-dimensional
projections. Tours can be considered to extend the dimensionality of
visualization, which is important as data and models exist in more than
3D. The package \CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) provides a platform
for generating tours. It can produce a variety of tours, each paired
with a variety of possible displays. A user can make a grand, guided,
little, local or frozen tour, and display the resulting projected data
as a scatterplot, density plot, histogram, or even as Chernoff faces if
the projection dimension is more than 3.

This work adds a manual tour to the collection. The manual tour was
described in Cook and Buja
(\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user to
control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore their sensitivity to the structure within the projection. As
manual tours operate on only one variable at a time, they are
particularly useful once a feature of interest has been identified.

One way to identify ``interesting'' features is with the use of a guided
tour (Cook et al. \protect\hyperlink{ref-cook_grand_1995}{1995}). Guided
tours select a very specific path, which approaches a projection that
optimizes an objective function. The optimization used to guide the tour
is simulated annealing (Kirkpatrick, Gelatt, and Vecchi
\protect\hyperlink{ref-kirkpatrick_optimization_1983}{1983}). The direct
optimization of a function allows guided tours to rapidly identify
interesting projection features given the relatively large
parameter-space. After a projection of interest is identified, an
analyst can then use the ``finer brush'' of the manual tour to control
the contributions of individual variables to explore the sensitivity
they have on the structure visible in the projection.

The paper is organized as follows. Section \ref{sec:algorithm} describes
the algorithm used to perform a radial manual tour as implemented in the
package \CRANpkg{spinifex}. Section \ref{sec:display} explains how to
generate an animation of the manual tour using the animation frameworks
offered by \CRANpkg{plotly} (Sievert
\protect\hyperlink{ref-sievert_plotly_2018}{2018}) and
\CRANpkg{gganimate} (Pedersen and Robinson
\protect\hyperlink{ref-pedersen_gganimate:_2019}{2019}). Package
functionality and code usage following the order applied in the
algorithm follows in section \ref{sec:usage}. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to multivariate data collected on high energy physics
experiments (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}). Section
\ref{sec:discussion} summarizes this paper and discusses potential
future directions.

@@

The algorithm to conduct a manual tour interactively, by recording
mouse/cursor motion, is described in detail in Cook and Buja
(\protect\hyperlink{ref-cook_manual_1997}{1997}). Movement can be in any
direction and magnitude, but it can also be constrained in several ways:

@@

The flea data (Lubischew
(\protect\hyperlink{ref-lubischew_use_1962}{1962})), available in the
\pkg{tourr} package, is used to illustrate the algorithm. The data
contains 74 observations on 6 variables, which are physical measurements
made on flea beetles. Each observation belongs to one of three species.

An initial 2D projection basis must be provided. A suggested way to
start is to identify an interesting projection using a projection
pursuit guided tour. Here the holes index is used to find a 2D
projection of the flea data, which shows three separated species groups.
Figure \ref{fig:step0} shows the initial projection of the data. The
left panel displays the projection basis (\(\textbf{B}\)) and can be
used as a visual guide of the magnitude and direction that each variable
contributes to the projection. The right panel shows the projected data,
\(\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}\). The
color and shape of points are mapped to the flea species. This plot is
made using the \code{view\_basis()} function in \pkg{spinifex}, which
generates a \CRANpkg{ggplot2} (Wickham
\protect\hyperlink{ref-wickham_ggplot2:_2016}{2016}) object.

@@

Rotations in 3D can be defined by the axes they pivot on. Rotation
within the projection, \(\theta\), is rotation around the \(Z\) axis.
Out-of-projection rotation, \(\phi\), is the rotation around an axis on
the \(XY\) plane, \(\textbf{U}\), orthogonal to \(\textbf{e}\). Given
these axes, the rotation matrix, \(\textbf{R}\) can be written as
follows, using Rodrigues' rotation formula (originally published in
Rodrigues (\protect\hyperlink{ref-rodrigues_lois_1840}{1840})):

@@

\hypertarget{sec:usage}{%
\section{Package structure and functionality}\label{sec:usage}}

@@

\begin{Schunk}
\begin{Sinput}
install.package("spinifex")
library("spinifex")
\end{Sinput}
\end{Schunk}

\noindent Also see the vignette for examples of usage:

\begin{verbatim}
vignette("spinifex_vignette")
\end{verbatim}

@@

\begin{verbatim}
remotes::install_github("nspyrison/spinifex") 
\end{verbatim}

@@

\caption{\label{tab:functionsTable}Summary of available functions.}
\centering
\begin{tabular}[t]{lll}
\toprule
Type & Function & Description\\
\midrule
construction & create\_manip\_space & forms the 3D space of rotation\\
construction & rotate\_manip\_space & performs 3D rotation\\
construction & manual\_tour & generates sequence of 2D frames\\
 &  & \\
render & array2df & turn the tour path array into long form, for plotting\\
render & render\_ & render long form as a ggplot2 objection for animation\\
render & render\_plotly & render the animation as a plotly object (default)\\
render & render\_gganimate & render the animation as a gganimate object\\
 &  & \\
animation & play\_tour\_path & composite function animating the specified tour path\\
animation & play\_manual\_tour & composite function animating the specified manual tour\\
 &  & \\
specialty & view\_basis & display the reference frame of a given basis\\
specialty & view\_manip\_space & illustrative display of any manip space\\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{usage}{%
\subsection{Usage}\label{usage}}

@@

\begin{Schunk}
\begin{Sinput}
library(spinifex)
f_data  <- tourr::rescale(flea[, 1:6])                    ## standardized flea data
f_path  <- save_history(f_data, guided_tour(holes()))     ## guided tour path, holes index
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2) ## optimal projection
f_cat   <- factor(flea$species)                           ## categorical var
f_mvar  <- 5                                              ## set manip var number
angle_speed <- .26                                        ## anglular dist between slides
play_manual_tour(data = f_data,                           ## generates animation
                 basis = f_basis,                         ##   plotly obj (default)
                 manip_var = f_mvar, 
                 angle = angle_speed,
                 col = f_cat,
                 pch = f_cat)
\end{Sinput}
\end{Schunk}

\noindent The \code{play\_manual\_tour()} function is a composite
function for user convenience that does these operations:

\begin{Schunk}
\begin{Sinput}
manual_tour()
array2df()
render_plotly()
\end{Sinput}
\end{Schunk}

\noindent This will generate an html animation using plotly. Switching
the renderer to \code{gganimate} is an option, which will also create an
html animation or an animated gif. Each of these formats allows for the
animation to be made available on a web site, or directly visible in an
html formatted document.

@@

\begin{Schunk}
\begin{Sinput}
view_basis(basis = f_basis, 
           data = f_data,
           labels = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
view_manip_space(basis = f_basis, 
                 manip_var = f_mvar, 
                 labels = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

Wang et al. (\protect\hyperlink{ref-wang_mapping_2018}{2018}) introduces
a new tool, PDFSense, to visualize the sensitivity of hadronic
experiments to nucleon structure. The parameter-space of these
experiments lies in 56 dimensions, \(\delta \in \mathbb{R}^{56}\), and
are visualized as 3D subspaces of the 10 first principal components in
linear (PCA) and non-linear (t-SNE) embeddings.

Cook, Laa, and Valencia
(\protect\hyperlink{ref-cook_dynamical_2018}{2018}) illustrates how to
learn more about the structures using a grand tour. Tours can better
resolve the shape of clusters, intra-cluster detail, and better outlier
detection than PDFSense \& TFEP (TensorFlow embedded projections) or
traditional static embeddings. This example builds from here,
illustrating how the manual tour can be used to examine the sensitivity
of structure in a projection to different parameters. The specific 2D
projections passed to the manual tour were provided in their work.

@@

The jet cluster resides in a smaller dimensionality than the full set of
experiments with four principal components explaining 95\% of the
variation in the cluster (Cook, Laa, and Valencia
\protect\hyperlink{ref-cook_dynamical_2018}{2018}). The data within this
4D embedding is further subsetted, to ATLAS7old and ATLAS7new, to focus
on two groups that occupy different parts of the subspace. Radial manual
tours controlling contributions from PC4 and PC3 are shown in Figures
\ref{fig:JetClusterGood} and \ref{fig:JetClusterBad}, respectively. The
difference in shape can be interpreted as the experiments probing
different phase-spaces. Back-transforming the principal components to
the original variables can be done for a more detailed interpretation.

@@

Following Cook, Laa, and Valencia
(\protect\hyperlink{ref-cook_dynamical_2018}{2018}), to explore the DIS
cluster, PCA is recomputed and the first six principal components,
explaining 48\% of the full sample variation, are used. The
contributions of PC6 and PC2 are explored in Figures
\ref{fig:DISclusterGood} and \ref{fig:DISclusterBad}, respectively.
Three experiments are examined: DIS HERA1+2 (green), dimuon SIDIS
(purple), and charm SIDIS (orange).

@@

Dynamic linear projections of numeric multivariate data, tours, play an
important role in data visualization; they extend the dimensionality of
visuals to peek into high-dimensional data and parameter spaces. This
research has taken the manual tour algorithm, specifically the radial
rotation, used in GGobi (Swayne et al.
\protect\hyperlink{ref-swayne_ggobi:_2003}{2003}) to interactively
rotate a variable into or out of a 2D projection, and modified it to
create an animation that performs the same task. It is most useful for
examining the importance of variables, and how the structure in the
projection is sensitive or not to specific variables. This functionality
available in package \pkg{spinifex}. The work complements the methods
available in the \pkg{tourr} package.

@@

The manual tour is constrained in the sense that the effect of one
variable is dependent on the contributions of other variables in the
manip space. However, this can be useful to simplify a projection by
removing variables without affecting the visible structure. Defining a
manual rotation in high dimensions is possible using Givens rotations
and Householder reflections as outlined in Buja et al.
(\protect\hyperlink{ref-buja_computational_2005}{2005}). This would
provide more flexible manual rotation, but more difficult for a user
because they have the choice (too much choice) of which directions to
move.

@@

Having a graphical user interface would be useful for making it easier
and more accessible to a general audience. This is possible to implement
using \CRANpkg{shiny} (Chang et al.
\protect\hyperlink{ref-chang_shiny:_2018}{2018}). The primary purposes
of the interface would be to allow the user to interactively change the
manip variable easily, and the interpolation step for more or less
detailed views.

@@

This article was created in R, using \CRANpkg{knitr} (Xie
\protect\hyperlink{ref-stodden_knitr:_2014}{2014}) and
\CRANpkg{rmarkdown} (Xie, Allaire, and Grolemund
\protect\hyperlink{ref-xie_r_2018}{2018}), with code generating the
examples inline. The source files for this article be found at
\href{https://github.com/nspyrison/spinifex_paper/}{github.com/nspyrison/spinifex\_paper/}.
The source code for the \pkg{spinifex} package can be found at
\href{https://github.com/nspyrison/spinifex/}{github.com/nspyrison/spinifex/}.

\hypertarget{bibliography}{%
\section*{Bibliography}\label{bibliography}}
\addcontentsline{toc}{section}{Bibliography}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-asimov_grand_1985}{}%
Asimov, Daniel. 1985. ``The Grand Tour: A Tool for Viewing
Multidimensional Data.'' \emph{SIAM Journal on Scientific and
Statistical Computing} 6 (1): 128--43.
\url{https://doi.org/https://doi.org/10.1137/0906011}.

\leavevmode\hypertarget{ref-buja_computational_2005}{}%
Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
``Computational Methods for High-Dimensional Rotations in Data
Visualization.'' In \emph{Handbook of Statistics}, 24:391--413.
Elsevier. \url{https://doi.org/10.1016/S0169-7161(04)24014-7}.

\leavevmode\hypertarget{ref-chang_shiny:_2018}{}%
Chang, Winston, Joe Cheng, J. J. Allaire, Yihui Xie, and Jonathan
McPherson. 2018. \emph{Shiny: Web Application Framework for R}.
\url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-cook_manual_1997}{}%
Cook, Dianne, and Andreas Buja. 1997. ``Manual Controls for
High-Dimensional Data Projections.'' \emph{Journal of Computational and
Graphical Statistics} 6 (4): 464--80.
\url{https://doi.org/10.2307/1390747}.

\leavevmode\hypertarget{ref-cook_grand_1995}{}%
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
``Grand Tour and Projection Pursuit.'' \emph{Journal of Computational
and Graphical Statistics} 4 (3): 155.
\url{https://doi.org/10.2307/1390844}.

\leavevmode\hypertarget{ref-cook_dynamical_2018}{}%
Cook, Dianne, Ursula Laa, and German Valencia. 2018. ``Dynamical
Projections for the Visualization of PDFSense Data.'' \emph{Eur. Phys.
J. C} 78 (9): 742. \url{https://doi.org/10.1140/epjc/s10052-018-6205-2}.

\leavevmode\hypertarget{ref-kirkpatrick_optimization_1983}{}%
Kirkpatrick, Scott, C. Daniel Gelatt, and Mario P. Vecchi. 1983.
``Optimization by Simulated Annealing.'' \emph{Science} 220 (4598):
671--80. \url{https://doi.org/10.1126/science.220.4598.671}.

\leavevmode\hypertarget{ref-lubischew_use_1962}{}%
Lubischew, Alexander A. 1962. ``On the Use of Discriminant Functions in
Taxonomy.'' \emph{Biometrics}, 455--77.
\url{https://doi.org/10.2307/2527894}.

\leavevmode\hypertarget{ref-pedersen_gganimate:_2019}{}%
Pedersen, Thomas Lin, and David Robinson. 2019. \emph{Gganimate: A
Grammar of Animated Graphics}.
\url{http://github.com/thomasp85/gganimate}.

\leavevmode\hypertarget{ref-rodrigues_lois_1840}{}%
Rodrigues, Olinde. 1840. \emph{Des Lois Géométriques Qui Régissent Les
Déplacements d'un Système Solide Dans L'espace: Et de La Variation Des
Cordonnées Provenant de Ces Déplacements Considérés Indépendamment Des
Causes Qui Peuvent Les Produire}.

\leavevmode\hypertarget{ref-sievert_plotly_2018}{}%
Sievert, Carson. 2018. \emph{Plotly for R}.
\url{https://plotly-book.cpsievert.me}.

\leavevmode\hypertarget{ref-swayne_ggobi:_2003}{}%
Swayne, Deborah F, Duncan Temple Lang, Andreas Buja, and Dianne Cook.
2003. ``GGobi: Evolving from XGobi into an Extensible Framework for
Interactive Data Visualization.'' \emph{Computational Statistics \& Data
Analysis}, Data Visualization, 43 (4): 423--44.
\url{https://doi.org/10.1016/S0167-9473(02)00286-4}.

\leavevmode\hypertarget{ref-wang_mapping_2018}{}%
Wang, Bo-Ting, T. J. Hobbs, Sean Doyle, Jun Gao, Tie-Jiun Hou, Pavel M.
Nadolsky, and Fredrick I. Olness. 2018. ``Mapping the Sensitivity of
Hadronic Experiments to Nucleon Structure.'' \emph{Physical Review D} 98
(9): 094030. \url{https://doi.org/10.1103/PhysRevD.98.094030}.

\leavevmode\hypertarget{ref-wickham_ggplot2:_2016}{}%
Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data
Analysis}. Springer-Verlag New York. \url{http://ggplot2.org}.

\leavevmode\hypertarget{ref-wickham_tourr_2011}{}%
Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
``\textbf{Tourr} : An \emph{R} Package for Exploring Multivariate Data
with Projections.'' \emph{Journal of Statistical Software} 40 (2).
\url{https://doi.org/10.18637/jss.v040.i02}.

\leavevmode\hypertarget{ref-stodden_knitr:_2014}{}%
Xie, Yihui. 2014. ``Knitr: A Comprehensive Tool for Reproducible
Research in R.'' In \emph{Implementing Reproducible Computational
Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
Peng. Chapman; Hall/CRC.
\url{http://www.crcpress.com/product/isbn/9781466561595}.

\leavevmode\hypertarget{ref-xie_r_2018}{}%
Xie, Yihui, J. J. Allaire, and Garrett Grolemund. 2018. \emph{R
Markdown: The Definitive Guide}. Boca Raton, Florida: Chapman; Hall/CRC.
\url{https://bookdown.org/yihui/rmarkdown}.

>>>>

@@

Exploring multivariate spaces is a challenging task, increasingly so as
dimensionality increases. Traditionally, static low-dimensional
projections are used to display multivariate data in two dimensions
including principal component analysis, linear discriminant spaces or
projection pursuit. These are useful for finding relationships between
multiple variables, but they are limited because they show only a
glimpse of the high-dimensional space. An alternative approach is to use
a tour \citep{asimov_grand_1985} of dynamic linear projections to look
at many different low-dimensional projections. Tours can be considered
to extend the dimensionality of visualization, which is important as
data and models exist in more than 3D. The package \CRANpkg{tourr}
\citep{wickham_tourr:_2011} provides a platform for generating tours. It
can produce a variety of tours, each paired with a variety of possible
displays. A user can make a grand, guided, little, local or frozen tour,
and display the resulting projected data as a scatterplot, density plot,
histogram, or even as Chernoff faces if the projection dimension is more
than 3.

This work adds a manual tour to the collection. The manual tour was
described in \citet{cook_manual_1997} and allows a user to control the
projection coefficients of a selected variable in a 2D projection. The
manipulation of these coefficients allows the analyst to explore their
sensitivity to the structure within the projection. As manual tours
operate on only one variable at a time, they are particularly useful
once a feature of interest has been identified.

One way to identify ``interesting'' features is with the use of a guided
tour \citep{cook_grand_1995}. Guided tours select a very specific path,
which approaches a projection that optimizes an objective function. The
optimization used to guide the tour is simulated annealing
\citep{kirkpatrick_optimization_1983}. The direct optimization of a
function allows guided tours to rapidly identify interesting projection
features given the relatively large parameter-space. After a projection
of interest is identified, an analyst can then use the ``finer brush''
of the manual tour to control the contributions of individual variables
to explore the sensitivity they have on the structure visible in the
projection.

The paper is organized as follows. Section \ref{sec:algorithm} describes
the algorithm used to perform a radial manual tour as implemented in the
package \CRANpkg{spinifex}. Section \ref{sec:display} explains how to
generate an animation of the manual tour using the animation frameworks
offered by \CRANpkg{plotly} \citep{sievert_interactive_2020} and
\CRANpkg{gganimate} \citep{pedersen_gganimate_2020}. Package
functionality and code usage following the order applied in the
algorithm follows in section \ref{sec:usage}. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to multivariate data collected on high energy physics
experiments \citep{wang_mapping_2018}. Section \ref{sec:discussion}
summarizes this paper and discusses potential future directions.

@@

The algorithm to conduct a manual tour interactively, by recording
mouse/cursor motion, is described in detail in \citet{cook_manual_1997}.
Movement can be in any direction and magnitude, but it can also be
constrained in several ways:

@@

The flea data (\citet{lubischew_use_1962}), available in the \pkg{tourr}
package, is used to illustrate the algorithm. The data contains 74
observations on 6 variables, which are physical measurements made on
flea beetles. Each observation belongs to one of three species.

An initial 2D projection basis must be provided. A suggested way to
start is to identify an interesting projection using a projection
pursuit guided tour. Here the holes index is used to find a 2D
projection of the flea data, which shows three separated species groups.
Figure \ref{fig:step0} shows the initial projection of the data. The
left panel displays the projection basis (\(\textbf{B}\)) and can be
used as a visual guide of the magnitude and direction that each variable
contributes to the projection. The right panel shows the projected data,
\(\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}\). The
color and shape of points are mapped to the flea species. This plot is
made using the \code{view\_basis()} function in \pkg{spinifex}, which
generates a \CRANpkg{ggplot2} \citep{wickham_ggplot2_2016} object.

@@

Rotations in 3D can be defined by the axes they pivot on. Rotation
within the projection, \(\theta\), is rotation around the \(Z\) axis.
Out-of-projection rotation, \(\phi\), is the rotation around an axis on
the \(XY\) plane, \(\textbf{U}\), orthogonal to \(\textbf{e}\). Given
these axes, the rotation matrix, \(\textbf{R}\) can be written as
follows, using Rodrigues' rotation formula (originally published in
\citet{rodrigues_lois_1840}):

@@

\hypertarget{package-structure-and-functionality}{%
\section{Package structure and
functionality}\label{package-structure-and-functionality}}

@@

\begin{Schunk}
\begin{Sinput}
## Install from CRAN
install.package("spinifex")
## Load into session
library("spinifex")
\end{Sinput}
\end{Schunk}

\noindent Also see the shiny app for understandign and the vignette for
basic usage:

\begin{Schunk}
\begin{Sinput}
## Shiny app for visualizing basic application
run_app("intro")
## View the code vignette
vignette("spinifex_vignette")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
## Optionally install latest developmention version from GitHub
remotes::install_github("nspyrison/spinifex")
\end{Sinput}
\end{Schunk}

@@

\caption{\label{tab:functionsTable}Summary of available functions.}
\centering
\begin{tabular}[t]{lll}
\toprule
Type & Function & Description\\
\midrule
construction & create\_manip\_space & forms the 3D space of rotation\\
construction & rotate\_manip\_space & performs 3D rotation\\
construction & manual\_tour & generates sequence of 2D frames\\
 &  & \\
render & array2df & turn the tour path array into long form, for plotting\\
render & render\_ & render long form as a ggplot2 objection for animation\\
render & render\_plotly & render the animation as a plotly object (default)\\
render & render\_gganimate & render the animation as a gganimate object\\
 &  & \\
animation & play\_tour\_path & composite function animating the specified tour path\\
animation & play\_manual\_tour & composite function animating the specified manual tour\\
 &  & \\
specialty & oblique\_basis & table of the rotated basis and manip space\\
specialty & oblique\_frame & display the reference axes of a given basis\\
specialty & view\_manip\_space & illustrative display of any manip space\\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{sec:usage}{%
\subsection{Usage}\label{sec:usage}}

@@

\begin{Schunk}
\begin{Sinput}
library(spinifex)
## Standardized flea data
f_data <- tourr::rescale(flea[, 1:6])
## Guided tour path, holes index
f_path <- save_history(f_data, guided_tour(holes()))
## Local extrema found
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
## Categorical class variable
f_cat <- factor(flea$species)
## Manip var, number of the variable to alter
f_mvar <- 5
## Anglular dist between frames (radians)
step_size <- .26
## Render and play animate, as plotly object by default
play_manual_tour(data = f_data,
                 basis = f_basis,
                 manip_var = f_mvar, 
                 angle = step_size,
                 col = f_cat,
                 pch = f_cat)
\end{Sinput}
\end{Schunk}

\noindent The \code{play\_manual\_tour()} function is a composite
function handling interaction between \code{manual\_tour()},
\code{array2df()}, and \code{render\_plotly()}. This will generate an
html animation using plotly. Switching the renderer to
\code{render\_gganimate()} alternatively creates an animated gif. Each
of these formats allows for the animation to be made available on a web
site, or directly visible in an html formatted document.

@@

\begin{Schunk}
\begin{Sinput}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_cat,
              shape = f_cat)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
## Displays the projection plane and manipulation space for the 
view_manip_space(basis = f_basis, 
                 manip_var = f_mvar, 
                 lab = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

\citet{wang_mapping_2018} introduces a new tool, PDFSense, to visualize
the sensitivity of hadronic experiments to nucleon structure. The
parameter-space of these experiments lies in 56 dimensions,
\(\delta \in \mathbb{R}^{56}\), and are visualized as 3D subspaces of
the 10 first principal components in linear (PCA) and non-linear (t-SNE)
embeddings.

\citet{cook_dynamical_2018} illustrates how to learn more about the
structures using a grand tour. Tours can better resolve the shape of
clusters, intra-cluster detail, and better outlier detection than
PDFSense \& TFEP (TensorFlow embedded projections) or traditional static
embeddings. This example builds from here, illustrating how the manual
tour can be used to examine the sensitivity of structure in a projection
to different parameters. The specific 2D projections passed to the
manual tour were provided in their work.

@@

The jet cluster resides in a smaller dimensionality than the full set of
experiments with four principal components explaining 95\% of the
variation in the cluster \citep{cook_dynamical_2018}. The data within
this 4D embedding is further subsetted, to ATLAS7old and ATLAS7new, to
focus on two groups that occupy different parts of the subspace. Radial
manual tours controlling contributions from PC4 and PC3 are shown in
Figures \ref{fig:JetClusterGood} and \ref{fig:JetClusterBad},
respectively. The difference in shape can be interpreted as the
experiments probing different phase-spaces. Back-transforming the
principal components to the original variables can be done for a more
detailed interpretation.

@@

Following \citet{cook_dynamical_2018}, to explore the DIS cluster, PCA
is recomputed and the first six principal components, explaining 48\% of
the full sample variation, are used. The contributions of PC6 and PC2
are explored in Figures \ref{fig:DISclusterGood} and
\ref{fig:DISclusterBad}, respectively. Three experiments are examined:
DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange).

@@

Dynamic linear projections of numeric multivariate data, tours, play an
important role in data visualization; they extend the dimensionality of
visuals to peek into high-dimensional data and parameter spaces. This
research has taken the manual tour algorithm, specifically the radial
rotation, used in GGobi \citep{swayne_ggobi:_2003} to interactively
rotate a variable into or out of a 2D projection, and modified it to
create an animation that performs the same task. It is most useful for
examining the importance of variables, and how the structure in the
projection is sensitive or not to specific variables. This functionality
available in package \pkg{spinifex}. The work complements the methods
available in the \pkg{tourr} package.

@@

The manual tour is constrained in the sense that the effect of one
variable is dependent on the contributions of other variables in the
manip space. However, this can be useful to simplify a projection by
removing variables without affecting the visible structure. Defining a
manual rotation in high dimensions is possible using Givens rotations
and Householder reflections as outlined in
\citet{buja_computational_2005}. This would provide more flexible manual
rotation, but more difficult for a user because they have the choice
(too much choice) of which directions to move.

@@

Having a graphical user interface would be useful for making it easier
and more accessible to a general audience. This is possible to implement
using \CRANpkg{shiny} \citep{chang_shiny_2020}. The primary purposes of
the interface would be to allow the user to interactively change the
manip variable easily, and the interpolation step for more or less
detailed views.

@@

This article was created in R, using \CRANpkg{knitr}
\citep{xie_knitr_2020} and \CRANpkg{rmarkdown}
\citep{allaire_rmarkdown_2020}, with code generating the examples
inline. The source files for this article be found at
\href{https://github.com/nspyrison/spinifex_paper/}{github.com/nspyrison/spinifex\_paper/}.
The source code for the \pkg{spinifex} package can be found at
\href{https://github.com/nspyrison/spinifex/}{github.com/nspyrison/spinifex/}.
